Software architectures for workflow integration in science gateways

### 2016-04-20 paper.tex
Section 1 is now readable, although not final. Section 2 was merged into Section 1.

\title{Architectures for workflow integration in science gateways}

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

@@

\paragraph{Context} The question of integrating workflow engines in
science gateways can be seen at various levels, corresponding to
various definitions of workflows. One level is the SHIWA level, where
it was considered that workflow engines are aware of the DCI (and 'D'
is important because of data transfers, proxies, etc). Another level
is to remove 'D' from the definition: a workflow engine becomes a
program that submits jobs, potentially only to local clusters. It
opens a whole new class of workflow engines that we used to consider
as "applications". For instance, in neuroinformatics: Nipype, PSOM,
but also FSL through the fslsub tool. A workflow engine is not
supposed to be aware of the science gateway.  A wide-array of workflow
engines are available with specificies coming from the application
domain, available tools, etc Their integration in science gateways
becomes critical. Several of the examples presented in the
paper will be taken from medical image analysis, in particular
neuroimaging.

\paragraph{Goal} this paper reviews and compares the architectures to
integrate workflow engines in science gateways.

\paragraph{Contributions}
\begin{itemize}
\item We evaluate architectures based on our experience with existing systems
\item We propose a new architecture
\end{itemize}

\section{Workflow engines and science gateways}

\todo{This section still needs to be finished}

@@

In the last decade, the e-Science workflow community has developed
workflow systems to help developers access distributed infrastructures
such as clusters, grids, clouds and web services, resulting in tools
such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE/gUSE~\cite{Kacsuk2012}\footnote{\todo{WS-PGRADE/gUSE is the
    science gateway, check what the engine is called now}}, etc. Such
workflow engines usually describe applications in a  language
that offers parallel data and control flow constructs, visual edition
features, links with domain-specific tool repositories, provenance
recording, etc.
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

@@

We are looking for software architectures that are able to integrate
e-Science and domain-specific workflow engines in science
gateways. For the sake of the analysis conducted in this paper, we
define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
software that submits interdependent computing \emph{tasks} to an
infrastructure (local server, cluster, grid or cloud) based on a
workflow description (a.k.a \emph{workflow}) and using input data that
may be files, data base entries, or simple parameter values. Although
simplistic, this definition covers both e-Science and domain-specific
engines. Some workflow engines, usually e-Science ones, may transfer
data across the infrastructure, and others, usually domain-specific
ones, may leave it to external processes. Workflows may be expressed
in any language, including high-level XML or JSON dialects such as
\texttt{Scufl} or \texttt{Hyperflow}, and low-level scripts such as
\texttt{bash}.  \todo{Mention other workflow definitions,
  e.g. abstract/concrete, etc?} \todo{Use ttt everywhere for technical
  terms (ssh, etc)?}

@@

Science gateways are ecosystems of software tools and services to
access data and tools deployed on distributed infrastructures. They
are used to share resources in a community and to provide performance
through facilitated access to storage and computing power. They are
usually accessible through a web interface that helps users manage
access rights, data transfers, task execution and authentication on
multiple computing and storage locations. Workflow engines are of
course part of this ecosystem as core components to implement and
execute applications.

@@

This paper builds on experience from the development and sustained
operations of CBRAIN (2008-now), the Virtual Imaging Platform
(2009-now), and on our involvement with the
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} European project
(2010-2012) that built the SHIWA Simulation Platform.

CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
open-source, web-based collaborative platform addressing the
challenges of data-heavy, computation-intensive neuroimaging
research. It offers transparent web-based access to remote data
sources, distributed computing sites, and an array of processing and
visualization tools, within a secure environment. CBRAIN is deployed
at a wide range of computing sites, including nine national HPC
centers in Canada, one in Korea, one in Germany, and on numerous local
servers. As September 2015, CBRAIN has served over 370 users across 50
cities in 19 countries who produced several high-impact results in the
field of neurosciences.

@@

\subsection{Infrastructures}

The infrastructure consists of the computing and storage resources
involved in the workflow execution, and the software services used to
access these resources. Infrastructures can be servers, clusters,
grids or clouds. Some workflow engines may assume specific
characteristics about the infrastructure, such as the presence of a
shared file system between the computing nodes.

@@

>>>>

\title{Software architectures for workflow integration in science gateways}

@@

\begin{keyword}
Workflow engines \sep science gateways \sep software architecture.
\end{keyword}

@@

Workflow engines are critical components of the ecosystem of tools and
services offered by science gateways to leverage distributed
infrastructures for the efficient execution of applications through
the web. Several software architectures can be adopted to integrate
workflow engines in science gateways, with important consequences on
the required development effort and on the characteristics of the
resulting system. This paper describes and compares such architectures
based on system-independent representations of their main components
and interactions.

Our analysis is fed by experience gathered from the development and
sustained operation of the CBRAIN~\cite{SHER-14} and
VIP~\cite{GLAT-13} science gateways for medical image analysis during
the past 7 years. It also builds on the lessons learned from several
science gateway and workflow-related projects, including the SHIWA
Simulation Platform~\cite{terstyanszky2014enabling}.

The remainder of this Section provides background information and
definitions about workflow engines and science gateways. In
Section~\ref{sec:architectures}, we describe 6 architectures in a
consistent framework based on the functional interactions between the
main software components involved. In Section~\ref{sec:evaluation}, we
introduce metrics to measure integration effort, robustness,
extensibility, scalability and other specific features related to the
architectures. We use these metrics to evaluate and compare the 6
architectures. The paper closes on a discussion and conclusion that
highlight architectural recommendations for science gateway and
workflow engine developers.

@@

In the last decade, the e-Science workflow community has developed
workflow systems to help developers access distributed infrastructures
such as clusters, grids, clouds and web services. These efforts
resulted in tools such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE/gUSE~\cite{Kacsuk2012}\footnote{\todo{WS-PGRADE/gUSE is the
    science gateway, check what the engine is called now}}, etc. Such
workflow engines usually describe applications in a specific language
that offers parallel data and control flow constructs, visual edition
features, links with domain-specific tool repositories, provenance
recording, among other features.
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

@@

For the sake of the analysis conducted in this paper, we
define a \emph{workflow engine} (also abbreviated \emph{engine}) as a
software that submits interdependent computing \emph{tasks} to an
infrastructure (local server, cluster, grid or cloud) based on a
workflow description (a.k.a \emph{workflow}) and using input data that
may be files, data base entries, or simple parameter values. Although
simplistic, this definition covers both e-Science and domain-specific
engines. Some workflow engines, usually e-Science ones, may transfer
data across the infrastructure, and others, usually domain-specific
ones, may leave it to external processes. Workflows may be expressed
in any language, including high-level XML or JSON dialects such as
\texttt{Scufl} or \texttt{Hyperflow}, and low-level scripts such as
\texttt{bash}.  \todo{Mention other workflow definitions,
  e.g. abstract/concrete, etc?} \todo{Use ttt everywhere for technical
  terms (ssh, etc)?}

\todo{The infrastructure consists of the computing and storage resources
involved in the workflow execution, and the software services used to
access these resources. Infrastructures can be servers, clusters,
grids or clouds. Some workflow engines may assume specific
characteristics about the infrastructure, such as the presence of a
shared file system between the computing nodes.
}

@@

Science gateways 
are used to share resources in a community and to provide performance
through facilitated access to storage and computing power. They are
usually accessible through a web interface that helps users manage
access rights, data transfers, task execution and authentication on
multiple computing and storage locations. Workflow engines are of
course part of this ecosystem as core components to implement and
execute applications.

@@

\todo{Transition needed}
CBRAIN\footnote{\url{http://cbrain.mcgill.ca}}~\cite{SHER-14} is an
open-source, web-based collaborative platform addressing the
challenges of data-heavy, computation-intensive neuroimaging
research. It offers transparent web-based access to remote data
sources, distributed computing sites, and an array of processing and
visualization tools, within a secure environment. CBRAIN is deployed
at a wide range of computing sites, including nine national HPC
centers in Canada, one in Korea, one in Germany, and on numerous local
servers. As September 2015, CBRAIN has served over 370 users across 50
cities in 19 countries who produced several high-impact results in the
field of neurosciences.

@@

@@

\label{sec:evaluation}
